---
title: "EDA"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
---

```{r setup, include=FALSE, message=FALSE}
library(auk)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(readr)
library(sf)
library(kableExtra)
library(leaflet)

# Set default figure options
knitr::opts_chunk$set(
  fig.width = 6,
  out.width = "90%"
)
```


```{r, include=FALSE, message=FALSE}
observations_selected = read_csv("data/observations_selected_clean.csv")
```


<br><br> 

## Season

**Does the season affect the activity of woodcocks?**

American woodcock is a migratory species in the eastern United States, so the chances of seeing them throughout the year vary. We will now look at the observations of woodcocks for each season.


```{r}
# Define colors for the seasons
season_colors <- c(
  "Winter" = "#56B4E9", 
  "Spring" = "#009E73",  
  "Summer" = "#F0E442",  
  "Fall" = "#E69F00"
)

# Function to assign colors based on seasons
season_color <- function(season) {
  unname(season_colors[as.character(season)])  # Use color specific to the season
}

# Render the interactive map
leaflet(data = observations_selected) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    ~longitude, ~latitude,  # Longitude and Latitude
    radius = 3,            # Marker size
    fillColor = ~season_color(season),  # Apply season colors
    color = "grey",        # Grey outline for all markers
    fillOpacity = 0.6,     # Set fill opacity for transparency
    stroke = TRUE,         # Add borders to markers
    weight = 0.5,          # Border thickness
    popup = ~paste("Season:", season, "<br>",
                   "Observation Count:", observation_count, "<br>",
                   "Time:", substr(time_observations_started, 1, 5), "<br>",  # Extract HH:MM
                   "Date:", observation_date)  # Add popup information
  ) %>%
  addLegend(
    position = "bottomright",
    colors = unname(season_colors),  # Colors for the legend
    labels = names(season_colors),  # Labels for the seasons
    title = "Season",
    opacity = 1
  )
```





:::: {style="display: flex;"}

::: {}

```{r, fig.align='center', fig.width=6.5, fig.height=4}
observations_selected <- observations_selected |> 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

season_summary <- observations_selected |> 
  group_by(season) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

ggplot(season_summary, aes(x = season, y = total_observation_count, fill = season)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Winter" = "#56B4E9", 
                               "Spring" = "#009E73", 
                               "Summer" = "#F0E442", 
                               "Fall" = "#E69F00")) +
  labs(
    title = "Observation Count of Woodcocks by Season in NY",
    x = "Season", 
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none",
       text = element_text(family = "Times", size = 14))
```


:::
  
:::{}

```{r, fig.align='center', fig.width=6.5, fig.height=4}
observations_selected <- observations_selected |> 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

season_summary <- observations_selected |> 
  group_by(season) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a table using kable for display
season_summary |> 
  kable(
    col.names = c("Season", "Total Observation Count"),
    caption = "Table: Total observation count of woodcocks by season in NY"
  )
```


:::
  
::::
  
Spring stands out with the highest observation count, likely due to their breeding behavior during this season, making them easier to detect by call. In contrast, observations are significantly fewer in winter, summer, and fall. 


<br><br> 






## Time of the day

**When do American woodcocks usually appear?**

Let us explore how the number of observations of woodcocks changes across the day in New York state. 


```{r}
# map

# Define colors for the times of day
time_colors <- c(
  "Morning" = "#FFD700",    # Gold for Morning
  "Afternoon" = "#FFA500",  # Orange for Afternoon
  "Evening" = "#1E90FF",    # Dodger Blue for Evening
  "Night" = "#4B0082"       # Indigo for Night
)

# Function to assign colors based on time of day
time_color <- function(time_of_day) {
  unname(time_colors[as.character(time_of_day)])  # Use color specific to the time of day
}

# Render the interactive map
leaflet(data = observations_selected) %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    ~longitude, ~latitude,  # Longitude and Latitude
    radius = 3,            # Marker size
    fillColor = ~time_color(time_of_day),  # Apply time of day colors
    color = "grey",        # Grey outline for all markers
    fillOpacity = 0.6,     # Set fill opacity for transparency
    stroke = TRUE,         # Add borders to markers
    weight = 0.5,          # Border thickness
    popup = ~paste("Time of Day:", time_of_day, "<br>",
                   "Observation Count:", observation_count, "<br>",
                   "Time:", substr(time_observations_started, 1, 5), "<br>",  # Extract HH:MM
                   "Date:", observation_date)  # Add popup information
  ) %>%
  addLegend(
    position = "bottomright",
    colors = unname(time_colors),  # Colors for the legend
    labels = names(time_colors),  # Labels for the times of day
    title = "Time of Day",
    opacity = 1
  )
```


:::: {style="display: flex;"}

::: {}


```{r, fig.align='center', fig.width=6.5, fig.height=4}

observations_selected <- observations_selected |> 
  mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))

# Summarize observation counts by time of day
time_summary <- observations_selected |> 
  group_by(time_of_day) |> 
  summarise(total_observation_count = n())

# Create the plot
ggplot(time_summary, aes(x = time_of_day, y = total_observation_count, fill = time_of_day)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "Morning" = "#FFD700", 
    "Afternoon" = "#FFA500", 
    "Evening" = "#1E90FF",  
    "Night" = "#4B0082" 
  )) +
  labs(
    title = "Observation Count by Time of Day",
    x = "Time of Day",
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


:::
  
:::{}


```{r, fig.align='center', fig.width=6.5, fig.height=4}
# table
# Ensure 'time_of_day' is a factor with a specific order
observations_selected <- observations_selected |> 
  mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))

# Summarize total observation count by time of day
time_summary <- observations_selected |> 
  group_by(time_of_day) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a table using kable for display
time_summary |> 
  kable(
    col.names = c("Time of Day", "Total Observation Count"),
    caption = "Table: Total observation count of woodcocks by time of day in NY"
  )
```

:::
  
::::
  

The graphs demonstrate clear patterns in the observation counts of American woodcocks across different times of the day in New York state: highest in the evening, followed by the morning. This pattern may be linked to their mating calls, which takes place right after sunset, making it easier for birders to encounter them during this time.


<br><br> 


**Are the difference significant?**

```{r}
kruskal_test <- kruskal.test(observation_count ~ time_of_day, data = observations_selected)

# Extract test results
kruskal_result <- data.frame(
  Statistic = round(kruskal_test$statistic, 3),
  Degrees_of_Freedom = kruskal_test$parameter,
  P_Value = format.pval(kruskal_test$p.value, digits = 3, scientific = TRUE)
)

# Save results as a nice table
kruskal_result |> 
  kbl(caption = "Kruskal-Wallis Test Results") |> 
  kable_classic(full_width = FALSE, html_font = "Times")
```

There is strong evidence that the observation counts differ significantly across the time periods (Morning, Afternoon, Evening, Night).


<br><br> 

## County

**Observation Counts by County**

American Woodcocks are found across New York State, but their activity levels vary by location. Habitat, environment, and observer efforts all play a role in where they are most commonly seen. Let us look at the plot below to see how they are distributed across the state by county.


:::: {style="display: flex;"}

::: {}

```{r}
# Summarize the data by county
county_summary <- observations_selected |> 
  filter(!is.na(county)) |> 
  group_by(county) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a bar plot
ggplot(county_summary, aes(x = reorder(county, -total_observation_count), y = total_observation_count, fill = total_observation_count)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
  labs(
    title = "Observation Counts by County",
    x = "County",
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        legend.position = "right",
        plot.title = element_text(size = 10, hjust = 0.5),  # Smaller title, centered
        axis.title.x = element_text(size = 8),  
        axis.title.y = element_text(size = 8),
        legend.text = element_text(size = 7)
        )
```

:::

:::{}



:::
::::


Tompkins County stands out with the highest number of observations, followed by several other counties with moderate counts. Most counties have relatively low observation numbers, indicating that woodcock activity or observation efforts may be concentrated in specific areas. Tompkins County hosts a huge eBird user population, as it is where Cornell Lab of Ornithology, the developing team of eBird, is located. It is worth noting that Tompkins County is the checklist leader of all counties in New York despite being 28th within the state by population and 46th by land area. Thus, the large number of observations in the county is more attributed to the number of observers rather than a high concentration of woodcocks.


<br><br> 

## Observer Behavior

The probability of spotting an American Woodcock differs across observers. Factors such as how long they spend looking, how far they travel, the time of day, and the methods they use all influence such probability. For example, observers who spend more time in areas where woodcocks are present are expected to have a better chance of seeing one. 

<br>

- <span style="font-size:18px;">**Time of day at which an observation was started**</span>

```{r, warning=FALSE, fig.height=3, fig.width=8}
observations_selected$start_hour <- as.numeric(format(as.POSIXct(observations_selected$time_observations_started, format = "%H:%M:%S"), "%H"))

hourly_counts <- 
  observations_selected |>
  count(start_hour)

ggplot(hourly_counts, aes(x = start_hour, y = n)) +
  geom_point(size = 2, color = "black") +
  geom_segment(aes(xend = start_hour, yend = 0), color = "black") +  
  labs(
    title = "Distribution of Observation Start Times",
    x = "Hours since midnight",
    y = "Frequency"
  ) +
  theme_minimal()
```

<br>

- <span style="font-size:18px;">**Observation duration hours**</span>

```{r, warning=FALSE, fig.height=3, fig.width=8}
observations_selected$duration_hours <- floor(observations_selected$duration_minutes / 30) * 0.5

# Count the number of observations for each 0.5-hour interval
duration_counts <- observations_selected |>
  filter(duration_hours < 7.5) |> 
  arrange(duration_hours) |> 
  count(duration_hours) 

# Adjust labels for the x-axis
duration_counts$duration_label <- ifelse(duration_counts$duration_hours == 0, "<0.5", as.character(duration_counts$duration_hours))

# Create the dot-and-line plot
ggplot(duration_counts, aes(x = duration_label, y = n)) +
  geom_point(size = 2, color = "black") +
  geom_segment(aes(xend = duration_label, yend = 0), color = "black") +  # Vertical lines
  labs(
    title = "Distribution of Observations by Duration (Nearest 0.5 Hours)",
    x = "Duration (Hours)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Tilt x-axis labels for readability
```


<br>



- <span style="font-size:18px;">**Effort Distance vs. Duration**</span>

To see how effort distance and time spent observing are connected, we created this scatter plot. It shows the relationship between distance and duration, with colors representing different times of the day. This helps highlight patterns in observation efforts.

:::: {style="display: flex;"}

::: {}

```{r, fig.width=20, fig.height=16, message=FALSE}

filtered_data <- observations_selected |> 
  filter(effort_distance_km <= 30,
         duration_minutes <= 800)
ggplot(filtered_data, aes(x = duration_minutes, y = effort_distance_km, color = time_of_day)) +
  geom_point(alpha = 0.7, size = 5) +  # Points with transparency and size for better readability
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear regression trend line
  scale_color_manual(
    values = c(
      "Morning" = "#FFD700",
      "Afternoon" = "#FFA500",
      "Evening" = "#1E90FF",
      "Night" = "#4B0082"
    ),
    name = "Time of Day"
  ) +
  labs(
    title = "Scatter Plot of Effort Distance vs. Duration by Time of Day",
    y = "Effort Distance (km)",
    x = "Duration (minutes)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 38),
    axis.text = element_text(size = 30),
    axis.title = element_text(size = 30),
    legend.text = element_text(size = 30),
    legend.title = element_text(size = 30),
    legend.position = "bottom"
  ) +
  guides(color = guide_legend(override.aes = list(size = 20)))  # Larger points in legend for better visibility
```

:::

:::{}

Morning and afternoon observations are more common, with a higher density of points at shorter durations and distances. Evening and night efforts are less frequent but show a slightly broader spread in distance and duration. 

:::
::::







