---
title: "statistical_analysis"
author: "Zhengyong Chen"
output: html_document
---

```{r setup, include=FALSE, message=FALSE}
library(auk)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(readr)
library(sf)
library(kableExtra)
library(leaflet)
```

Data copied from data_wrangling...

```{r}
amwo <- "data/ebd_US-NY_amewoo_201901_202401_smp_relSep-2024.txt"
amwo_samp <- "data/ebd_US-NY_amewoo_201901_202401_smp_relSep-2024_sampling.txt"
observations <- read_ebd(amwo)
checklists <- read_sampling(amwo_samp) # the 600MB file will take a while to load

checklists_shared <- read_sampling(amwo_samp, unique = FALSE)
checklists_shared |> 
  filter(!is.na(group_identifier)) |> 
  arrange(group_identifier) |> 
  select(sampling_event_identifier, group_identifier)

checklists_unique <- auk_unique(checklists_shared, checklists_only = TRUE)

checklists <- checklists |> 
  filter(all_species_reported)

observations <- observations |> 
  filter(all_species_reported)

observations <- observations %>%
  filter(checklist_id %in% checklists$checklist_id)
```


## Descriptive Statistics & Statistical Analysis 

...of American Woodcock in NY



**Select and add some useful variables from observations.**

```{r}
observations_selected <- 
  observations |>
  select(county, observation_count, county, locality_type, latitude, longitude, 
         observation_date, time_observations_started, protocol_type, duration_minutes,
         effort_distance_km, number_observers) |> 
  filter(observation_count != "X") |> 
  mutate(
    observation_count = as.numeric(observation_count),
    year = lubridate::year(as.Date(observation_date)),
    month = lubridate::month(as.Date(observation_date)),
    day = lubridate::day(as.Date(observation_date)),
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"),
    time_of_day = case_when(
      as.numeric(format(as.POSIXct(time_observations_started, format = "%H:%M:%S"), "%H")) >= 5 &
      as.numeric(format(as.POSIXct(time_observations_started, format = "%H:%M:%S"), "%H")) < 12 ~ "Morning",
      as.numeric(format(as.POSIXct(time_observations_started, format = "%H:%M:%S"), "%H")) >= 12 &
      as.numeric(format(as.POSIXct(time_observations_started, format = "%H:%M:%S"), "%H")) < 17 ~ "Afternoon",
      as.numeric(format(as.POSIXct(time_observations_started, format = "%H:%M:%S"), "%H")) >= 17 &
      as.numeric(format(as.POSIXct(time_observations_started, format = "%H:%M:%S"), "%H")) < 21 ~ "Evening",
      TRUE ~ "Night")
  )
```

**Does the season affect the activity of woodcocks?**

```{r}
observations_selected <- observations_selected |> 
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

season_summary <- observations_selected |> 
  group_by(season) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

ggplot(season_summary, aes(x = season, y = total_observation_count, fill = season)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Winter" = "#56B4E9", 
                               "Spring" = "#009E73", 
                               "Summer" = "#F0E442", 
                               "Fall" = "#E69F00")) +
  labs(
    title = "Observation Count of Woodcocks by Season in NY",
    x = "Season", 
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The total observation count in Spring in NY is significantly higher than in any other season...

**When do American woodcocks usually appear?**

```{r}
observations_selected <- observations_selected |> 
  mutate(time_of_day = factor(time_of_day, levels = c("Morning", "Afternoon", "Evening", "Night")))

# Summarize observation counts by time of day
time_summary <- observations_selected |> 
  group_by(time_of_day) |> 
  summarise(total_observation_count = n())

# Create the plot
ggplot(time_summary, aes(x = time_of_day, y = total_observation_count, fill = time_of_day)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c(
    "Morning" = "#FFD700", 
    "Afternoon" = "#FFA500", 
    "Evening" = "#1E90FF",  
    "Night" = "#4B0082" 
  )) +
  labs(
    title = "Observation Count by Time of Day",
    x = "Time of Day",
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Are the difference significant?

```{r}
kruskal_test <- kruskal.test(observation_count ~ time_of_day, data = observations_selected)

# Extract test results
kruskal_result <- data.frame(
  Statistic = round(kruskal_test$statistic, 3),
  Degrees_of_Freedom = kruskal_test$parameter,
  P_Value = format.pval(kruskal_test$p.value, digits = 3, scientific = TRUE)
)

# Save results as a nice table
kruskal_result |> 
  kbl(caption = "Kruskal-Wallis Test Results") |> 
  kable_classic(full_width = FALSE, html_font = "Times")
```

There is strong evidence that the observation counts differ significantly across the time periods (Morning, Afternoon, Evening, Night).


```{r}
# Summarize the data by county
county_summary <- observations_selected |> 
  filter(!is.na(county)) |> 
  group_by(county) |> 
  summarise(total_observation_count = sum(observation_count, na.rm = TRUE))

# Create a bar plot
ggplot(county_summary, aes(x = reorder(county, -total_observation_count), y = total_observation_count, fill = total_observation_count)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightblue", high = "darkblue", name = "Count") +
  labs(
    title = "Observation Counts by County",
    x = "County",
    y = "Total Observation Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "bottom")
```


```{r}
# Prepare the data
observations_selected_clean <- observations_selected |> 
  filter(!is.na(longitude) & !is.na(latitude))

# Define colors for four categories
category_colors <- c(
  "1" = "skyblue",  
  "2" = "orange", 
  "3" = "red",  
  ">3" = "purple"  
)

# Function to assign colors based on categories
custom_color <- function(count) {
  ifelse(
    count > 3, 
    category_colors[">3"],  # Orange for counts >3
    category_colors[as.character(count)]  # Use category-specific color for 1, 2, 3
  )
}

# Create an interactive map
leaflet(data = observations_selected_clean) |> 
  addTiles() |>   # Add default OpenStreetMap tiles
  addCircleMarkers(
    ~longitude, ~latitude,  # Longitude and Latitude
    radius = 4,            # Marker size
    fillColor = ~custom_color(observation_count),  # Apply category colors
    color = "grey",        # Grey outline for all markers
    fillOpacity = 0.6,     # Set fill opacity for transparency
    stroke = TRUE,         # Add borders to markers
    weight = 0.5,          # Border thickness
    popup = ~paste("Count:", observation_count, "<br>",
                   "Time:", substr(time_observations_started, 1, 5), "<br>",
                   "Date:", observation_date)  # Add popup information
  ) |> 
  addLegend(
    position = "bottomright",
    colors = unname(category_colors),  # Colors for the legend
    labels = names(category_colors),  # Labels for the categories
    title = "Count",
    opacity = 1
  ) |> 
  addControl(
    html = "NY Woodcocks Observation Map</h5>",  # Title text in HTML
    position = "topleft"  
  )
```

